On Misuses of the Kolmogorov--Smirnov Test for One-Sample Goodness-of-Fit

### 2023-05-05 manuscript/manuscript.tex
remade figures, table, and discussion

\usepackage[pagewise]{lineno}
%\linenumbers*[1]
% %% patches to make lineno work better with amsmath
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
 \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
 \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
 \renewenvironment{#1}%
 {\linenomath\csname old#1\endcsname}%
 {\csname oldend#1\endcsname\endlinenomath}}%
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
 \patchAmsMathEnvironmentForLineno{#1}%
 \patchAmsMathEnvironmentForLineno{#1*}}%

@@

Nonparametric bootstrap can also be used to approximate the null distribution
of the test statistic in~\eqref{eq:ks_fitted} except that a bias correction is 
needed
\citep{babu2004goodness}. Details of the nonparametric bootstrap procedure are
in the Appendix. The results from nonparametric bootstrap are similar to those
from the parametric bootstrap and, hence, are omitted.

@@

\jy{how about removing the panels for $-0.1$ and 0.1?}
\eds{seems like too few if those are removed. How about $\{-0.4,-0.2,0.2,0.4\}$
or $\{-0.30,-0.15,0.15,0.30\}$? Is it easy/quick to generate the plots?} 
\jy{Will get jobs done for $\{-0.4,-0.2,0.2,0.4\}$}

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{pp_s.pdf}
  \caption{PP-plots for the p-values obtained from KS test with serial
    dependence ignored and sample size $n = 200$ based on 1000
    replicates. The data have marginal distributions $N(8, 8)$ and
    $\Gamma(8, 1)$, with an AR(1) dependence structure on the standard normal
    scale. The AR coefficients were $\psi \in \{-0.3, -0.1,  0.1,  0.3\}$.
  }
  \label{fig:pp_s}
\end{figure}

The invalidity of the standard statistic for serially dependent data can be
illustrated by a simple simulation study. Consider a first-order autoregressive
or AR(1) model with marginal normal (or gamma) distribution. For each AR
coefficient $\psi \in \{-0.3, -0.1, 0.1, 0.3\}$, we generated 1000
series of length 200. For each series, we applied the KS statistic
in~\eqref{eq:ks_standard} to test $H_0$ that the $X_i$'s follow $N(8, 8)$ (or 
$\Gamma(8,1)$) marginally. The PP-plots of the p-values from $1000$ replicates 
for all the $\psi$ values are displayed in Figure~\ref{fig:pp_s}. As 
$\phi$ diverges from 0, the p-values look less likely to be generated from a 
standard uniform distributions. Positive $\psi$ values led to p-values with a 
distribution function higher than the standard uniform distribution function as 
there were more smaller p-values than there should be, and 
hence, %conservative 
liberal tests. Negative $\psi$ values led to p-values with a 
distribution function lower than the standard uniform distribution as there 
were more larger p-values than there should be, and hence, 
%liberal 
conservative tests.

@@

This process is semiparametric because the marginal distribution is
parametrically specified while the dependence structure of the stationary series
is only approximated by that of an ARMA process with normal distributions.
The closer the approximation is to the truth, the better
performance of the size of the KS test. The normal copula of the working ARMA
process provides a wide class of dependence structures.
If the true dependence indeed has an ARMA structure, this method is exact. When
the true dependence is not covered by the ARMA model, the working model may 
still give a reasonable approximation that can
be useful for practical purposes.

@@

\begin{enumerate}
\item
  Generate $Z_1^*, \ldots, Z_n^*$ from the working ARMA$(p, q)$ model with
  innovation variance $1 / \sigma^2$ such that the $Z_i$'s are marginally
  standard normal variables.
\item
  Form a bootstrap sample
  $X_i^* = F_{\hat\theta_n}^{-1} [\Phi(Z_i^*)]$, $i = 1, \ldots, n$, where
  $F_{\hat\theta_n}^{-1}$ is the quantile function of $F_{\hat\theta_n}$.
\item
  Fit $F_\theta$ to $X_1^*, \ldots, X_n^*$ to obtain estimate $\hat\theta_n^*$
  of $\theta$.
\item
  Obtain the empirical distribution function $F_n^*$ of the bootstrap sample
  $X_1^*, \ldots, X_n^*$.
\item
  Calculate bootstrap KS statistic
  \[
    D_n^* = \sup_x \lvert F_n^* (x) - F_{\hat\theta_n}^* (x) \rvert.
  \]
  \eds{check placement of * in second F}
  \jy{the start should be with hat theta from step 3}
\item
  Repeat the previous steps a large number $B$ times and use the empirical
  distribution of the $B$ test statistics to approximate
  the null distribution of the observed statistic.
\end{enumerate}
The p-value of $D_n$ is approximated by the proportion of the $D_n^*$
statistics that are greater than or equal to~$D_n$. Compared to the algorithm in
the last section, here we use $F_{\hat\theta_n}$ and $F_{\hat\theta_n}^*$ in
place of the known~$F$. \eds{check placement of * in second F}

@@

Figure~\ref{fig:pp_ssf} shows the PP-plots resulting p-valuess of the
procedure performed on data generated from same simulation settings as in
Section~\ref{sec:dependence}. For comparison, we also displayed the PP-plot of
a naive bootstrap procedure that adjusts for the unspecified parameters as in
Section~\ref{sec:fitted} but not for the serial dependence. 
The super naive approach that does not adjust for fitted parameters nor serial
dependence would have double sin, so the results are not reported. 
The p-values from
naively fitting parameters but not adjusting for dependence clearly deviates
from the standard uniform distribution except in the MA(1) case.
\jy{needs an explanation ...}
\eds{if we had used a different MA parameter other than 0.8, would we see  
similar behavior?}  \jy{yes. I have more results for other MA coefficients to
  show when we meet.}
Similar to Section~\ref{sec:dependence}, the working normal copula approach
gives p-values with the desired standard uniform distribution regardless of the
true dependence.

\begin{table}[tbp]
  \label{tab:power}
  \caption{Power of rejecting the null hypothesis of normal and gamma
    distribution at significance level 0.05 when the true distribution was
    $\Gamma(8, 1)$ and $N(8, 8)$, respectively. }
\centering
\begin{tabular}{llrrrr}
  \toprule
  &  & \multicolumn{2}{c}{$H_0:$ normal} &
  \multicolumn{2}{c}{$H_0$: gamma}\\
  \cmidrule(lr){3-4}\cmidrule(lr){5-6}
  Dependence & Method & $n = 100$ & $n = 200$ & $n = 100$ & $n = 200$\\
  \midrule
ARMA & naive & 0.40 & 0.66 & 0.52 & 0.78 \\ 
   & copula & 0.33 & 0.56 & 0.44 & 0.72 \\ 
  AR & naive & 0.41 & 0.66 & 0.46 & 0.70 \\ 
   & copula & 0.26 & 0.45 & 0.31 & 0.54 \\ 
  MA & naive & 0.37 & 0.73 & 0.51 & 0.82 \\ 
   & copula & 0.34 & 0.69 & 0.49 & 0.80 \\ 
   \bottomrule
\end{tabular}
\end{table}

Finally, we report a small study of the power of the one-sample KS test with
unspecified parameters and serially dependent data. Using the three dependence
models characterized by ARMA(1, 1), AR(2), and MA(1) in the simulation settings
in the last section, we generated serially
dependent data with marginal distributions $N(8, 8)$ left truncated by zero and
$\Gamma(8, 1)$. 
Then, for data from $\Gamma(8, 1)$, we tested $H_0$ that the
data has marginally a normal distribution; for data from $N(8, 8)$ left
truncated by zero, we tested $H_0$ that the data has marginally a gamma
distribution;  In both cases, no parameters were specified. Two sample sizes
were considered, $n \in \{100, 200\}$. Table~\ref{tab:power} summarizes the
power of rejecting the null hypothesis at significance level~0.05. The tests
based on a working copula have substantial power in rejecting the null
hypothesis and the power increases as sample size increases. The power from the
naive tests where the serial dependence is ignored, although the unspecified
parameters are accounted for (i.e., procedure from 
Section~\ref{sec:dependence}), appear to 
have higher power than the proposed tests, but this is not reliable since they 
do not hold their sizes as demonstrated in the Figures.

@@

The KS test has base assumptions that the hypothesized distribution is
completely specified and the data is independent. When these assumptions are
violated, the test is no longer accurate and remedies must be performed. In the
case of fitted parameters, parametric and non-parametric bootstrap can restore
the size of the test. A bias correction is required if the non-parametric form
is used \citep{babu2004goodness}. In the case of serially dependent data, 
a procedure using bootstrap with the normal copula determined by an
automatically selected ARMA model to account for dependency shows positive
results. When both assumptions are violated, i.e., where the data has serial 
dependence and parameters must be fitted, an adjusted copula procedure also 
shows positive results. The tests appear effective for a variety of families of 
distributions. The copula remedy is not a complete solution and has limitations. 
Regardless of the true dependence, we assume an ARMA($p,q$) dependence structure. 
Therefore, if the assumed dependence structure is a close approximation of the
truth, the approach can work. However, if the approximation is too far from the 
true dependence structure the approach does not completely remedy the issue. 
As well, tests were only performed with the normal copula. It is possible that 
other copulas could provide stronger results based on the true dependence of the 
data.

@@

Using the same notations in the text, the procedure is summarized as follows
\citep{babu2004goodness}.
\begin{enumerate}
\item
  Draw a random sample $X_1^*,...,X_n^*$ from the empirical distribution $F_n$
  with replacement
\item
  Fit $F_\theta$ to $X_1^*,...,X_n^*$ and obtain estimate 
	$\hat\theta_n^*$ of $\theta$.
\item
  Obtain the empirical distribution function $F_n^*$ of
  $X_1^*, \ldots,  X_n^*$. 
\item
  Calculate bootstrap KS statistic
  \[
    D_n^* = \sup_x | \sqrt{n}\left(F_n^* (x)- F_{\hat\theta_n^*}(x)\right) - B_n(x) |.
  \]
  where $B_{n}(x) = \sqrt{n}(F_{n}(x) - F_{\hat\theta_n}(x))$ is the known
  bias term \citep{babu2004goodness}.
\item
  Repeat the previous steps a large number $B$ times and use the empirical
  distribution of $D_n^*$ to approximate the null distribution of the observed
  statistic. 
\end{enumerate}
The p-value of $D_n$ is, again, approximated by the proportion of the $D_n^*$
statistics that are greater than or equal to~$D_n$. Note that the step~1 and
step~4 are different from the parametric bootstrap version.

@@

\end{document}
%%% LocalWords: 
%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% ispell-personal-dictionary: ".aspell.en.pws"
%%% fill-column: 80
%%% eval: (auto-fill-mode 1)
%%% End:

>>>>

\usepackage[]{lineno}
\linenumbers*[1]
% %% patches to make lineno work better with amsmath
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
 \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
 \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
 \renewenvironment{#1}%
 {\linenomath\csname old#1\endcsname}%
 {\csname oldend#1\endcsname\endlinenomath}}%
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
 \patchAmsMathEnvironmentForLineno{#1}%
 \patchAmsMathEnvironmentForLineno{#1*}}%

@@

Nonparametric bootstrap can also be used to approximate the null distribution
of the test statistic in~\eqref{eq:ks_fitted} except that a bias correction is 
needed \citep{babu2004goodness}; see details of the procedure in
in the Appendix. The results from nonparametric bootstrap are similar to those
from the parametric bootstrap and, hence, are omitted.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{pp_s.pdf}
  \caption{PP-plots for the p-values obtained from KS test with serial
    dependence ignored and sample size $n = 200$ based on 1000
    replicates. The data have marginal distributions $N(8, 8)$ and
    $\Gamma(8, 1)$, with an AR(1) dependence structure on the standard normal
    scale. The AR coefficients were $\psi \in \{-0.4, -0.2,  0.2,  0.4\}$.
  }
  \label{fig:pp_s}
\end{figure}

The invalidity of the standard statistic for serially dependent data can be
illustrated by a simple simulation study. Consider a first-order autoregressive
or AR(1) model with marginal normal (or gamma) distribution. For each AR
coefficient $\psi \in \{-0.4, -0.2, 0.2, 0.4\}$, we generated 1000
series of length 200. For each series, we applied the KS statistic
in~\eqref{eq:ks_standard} to test $H_0$ that the $X_i$'s follow $N(8, 8)$ (or 
$\Gamma(8,1)$) marginally. The PP-plots of the p-values from $1000$ replicates 
for all the $\psi$ values are displayed in Figure~\ref{fig:pp_s}. As 
$\phi$ diverges from 0, the p-values look less likely to be generated from a 
standard uniform distributions. Positive $\psi$ values led to p-values with a 
distribution function higher than the standard uniform distribution function as 
there were more smaller p-values than there should be, and hence, 
liberal tests. Negative $\psi$ values led to p-values with a 
distribution function lower than the standard uniform distribution as there 
were more larger p-values than there should be, and hence, conservative tests.

@@

This process is semiparametric because the dependence structure is specified by
a normal copula determined by an ARMA process with normally distributed errors.
The closer the approximation is to the truth, the better
performance of the size of the KS test. The normal copula of the working ARMA
process provides a wide class of dependence structures.
If the true dependence indeed has an ARMA structure, this method is exact. When
the true dependence is not covered by the ARMA model, the working model may 
still give a reasonable approximation that can
be useful for practical purposes.

@@

\begin{enumerate}
\item
  Generate $Z_1^*, \ldots, Z_n^*$ from the working ARMA$(p, q)$ model with
  innovation variance $1 / \sigma^2$ such that the $Z_i$'s are marginally
  standard normal variables.
\item
  Form a bootstrap sample
  $X_i^* = F_{\hat\theta_n}^{-1} [\Phi(Z_i^*)]$, $i = 1, \ldots, n$, where
  $F_{\hat\theta_n}^{-1}$ is the quantile function of $F_{\hat\theta_n}$.
\item
  Fit $F_\theta$ to $X_1^*, \ldots, X_n^*$ to obtain estimate $\hat\theta_n^*$
  of $\theta$.
\item
  Obtain the empirical distribution function $F_n^*$ of the bootstrap sample
  $X_1^*, \ldots, X_n^*$.
\item
  Calculate bootstrap KS statistic
  \[
    D_n^* = \sup_x \lvert F_n^* (x) - F_{\hat\theta_n^*} (x) \rvert.
  \]
\item
  Repeat the previous steps a large number $B$ times and use the empirical
  distribution of the $B$ test statistics to approximate
  the null distribution of the observed statistic.
\end{enumerate}
The p-value of $D_n$ is approximated by the proportion of the $D_n^*$
statistics that are greater than or equal to~$D_n$. Compared to the algorithm in
the last section, here we use $F_{\hat\theta_n^*}$ in place of the known~$F$.

@@

Figure~\ref{fig:pp_ssf} shows the PP-plots resulting p-values of the
procedure performed on data generated from same simulation settings as in
Section~\ref{sec:dependence}. For comparison, we also displayed the PP-plot of
a naive bootstrap procedure that adjusts for the unspecified parameters as in
Section~\ref{sec:fitted} but not for the serial dependence. 
The super naive approach that does not adjust for fitted parameters nor serial
dependence would have double sin, so the results are not reported. 
The p-values from
naively fitting parameters but not adjusting for dependence clearly deviates
from the standard uniform distribution except in the MA(1) case, where the
autocorrelation is nonzero with a short memory of only lag~1.
Similar to Section~\ref{sec:dependence}, the working normal copula approach
gives p-values with the desired standard uniform distribution regardless of the
true dependence.

\begin{table}[tbp]
  \label{tab:power}
  \caption{Power of rejecting the null hypothesis of normal and gamma
    distribution at significance level 0.05 when the true distribution was
    $\Gamma(8, 1)$ and $N(8, 8)$, respectively. }
\centering
\begin{tabular}{ll cccc}
  \toprule
  &  & \multicolumn{2}{c}{$H_0:$ normal} &
  \multicolumn{2}{c}{$H_0$: gamma}\\
  \cmidrule(lr){3-4}\cmidrule(lr){5-6}
  Dependence & Method & $n = 100$ & $n = 200$ & $n = 100$ & $n = 200$\\
  \midrule
ARMA & naive & 0.404 & 0.656 & 0.515 & 0.779 \\ 
 & copula & 0.330 & 0.556 & 0.442 & 0.722 \\ 
AR & naive & 0.411 & 0.655 & 0.455 & 0.705 \\ 
& copula & 0.261 & 0.448 & 0.313 & 0.544 \\ 
MA & naive & 0.374 & 0.728 & 0.514 & 0.817 \\ 
& copula & 0.343 & 0.693 & 0.487 & 0.797 \\ 
   \bottomrule
\end{tabular}
\end{table}

Finally, we report a small study of the power of the one-sample KS test with
unspecified parameters and serially dependent data. Using the three dependence
models characterized by ARMA(1, 1), AR(2), and MA(1) in the simulation settings
in the last section, we generated serially
dependent data with marginal distributions $N(8, 8)$ left truncated by zero and
$\Gamma(8, 1)$. 
Then, for data from $\Gamma(8, 1)$, we tested $H_0$ that the
data has marginally a normal distribution; for data from $N(8, 8)$ left
truncated by zero, we tested $H_0$ that the data has marginally a gamma
distribution;  In both cases, no parameters were specified. Two sample sizes
were considered, $n \in \{100, 200\}$. Table~\ref{tab:power} summarizes the
power of rejecting the null hypothesis at significance level~0.05 from
1000~replicates. The tests
based on a working copula have substantial power in rejecting the null
hypothesis and the power increases as sample size increases. The power from the
naive tests where the serial dependence is ignored, although the unspecified
parameters are accounted for (i.e., procedure from 
Section~\ref{sec:dependence}), appear to 
have higher power than the proposed tests, but this is not reliable since they 
do not hold their sizes as demonstrated in the Figure~\ref{fig:pp_ssf}.

@@

The three kinds of misuses of the KS test for one-sample goodness-of-fit that we
demonstrated here are commonly seen in practice. With the aid of computing tools
such as R, they can be easily demonstrated in classroom teaching. Practitioners
need to be aware of the assumptions of the standard one-sample KS test that the
hypothesized distribution is completely specified and that the data are
independent. When these assumptions are not met, naive applications of the test
are no longer accurate and remedies must be performed. In the
case of fitted parameters, parametric and nonparametric bootstrap can restore
the size of the test. The nonparametric version has long been developed too but
is lesser known \citep{babu2004goodness}. In the case of serially dependent data,
our semiparametric bootstrap procedure with a working dependence structure
controlled by an automatically selected ARMA model restore the size of the test
for a wide range of dependence structures.
When both assumptions are violated, i.e., where the data has serial 
dependence and the parameters must be fitted, a combination of the two
individual remedies shows good results.

The semiparametric bootstrap procedure with a working ARMA dependence structure
is not a complete solution and has limitations. When the true serial dependence
structure cannot be well approximated by an ARMA model with normally distributed
error terms, the remedy may not restore the size of the test. The working
dependence structure is essentially the normal copula uniquely determined by the
ARMA model with normal errors. Some copulas are quite different from the normal
copula. For example, extreme value copulas have tail dependence, which is not
provided by normal copulas.
Using a normal copula to approximate an extreme value copula may not capture the
true serial dependence structure that is needed to restore the size of the
test. A block version of the nonparametric bootstrap \citep{babu2004goodness},
which avoids specification of a working dependence structure, would be more
desirable.

@@

Using the same notations in the text, the nonparametric bootstrap procedure of
\citet{babu2004goodness} is summarized as follows
\citep{babu2004goodness}.
\begin{enumerate}
\item
  Draw a random sample $X_1^*,...,X_n^*$ from the empirical distribution $F_n$
  with replacement
\item
  Fit $F_\theta$ to $X_1^*,...,X_n^*$ and obtain estimate 
	$\hat\theta_n^*$ of $\theta$.
\item
  Obtain the empirical distribution function $F_n^*$ of
  $X_1^*, \ldots,  X_n^*$. 
\item
  Calculate bootstrap KS statistic
  \[
    D_n^* = \sup_x | \sqrt{n}\left(F_n^* (x)- F_{\hat\theta_n^*}(x)\right) - B_n(x) |.
  \]
  where $B_{n}(x) = \sqrt{n}(F_{n}(x) - F_{\hat\theta_n}(x))$ is the known
  bias term \citep{babu2004goodness}.
\item
  Repeat the previous steps a large number $B$ times and use the empirical
  distribution of $D_n^*$ to approximate the null distribution of the observed
  statistic. 
\end{enumerate}
The p-value of $D_n$ is, again, approximated by the proportion of the $D_n^*$
statistics that are greater than or equal to~$D_n$. Note that the step~1 and
step~4 are different from the parametric bootstrap version.

@@

\end{document}
%%% LocalWords: nonparametric semiparametric autocorrelation ARMA
%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% ispell-personal-dictionary: ".aspell.en.pws"
%%% fill-column: 80
%%% eval: (auto-fill-mode 1)
%%% End:
